The Two Dualities of Computation: Negative and Fractional Types

### 2012-02-28 rational.tex
spell check and notes on algebraic types

In a technical sense, this paper extends the language of isomorphisms
{{langRev}}, with duality. Unlike Linear logic \cite{Girard87tcs} and
other systems which have one notion of duality over additive and
multiplicative componenents, {{langRevEE}} has two notions of duality
-- an additive duality over the monoid {{(0, +)}} and a multiplicative
duality over the monoid {{(1, *)}}. Each axis of duality also give us
a function space and hence {{langRevEE}} has an additive function
space correspoding to a notion of control or backtracking and a
multiplicative function space corresponding to a notion of unification
or constraint satisfaction.

@@

The presentation here differs from \cite{infeffects} in the sense that
we have introduced the {{0}} type here, thereby completeing the monoid
{{(0,+)}} and we present a small-step operational semantics.

@@

We define observables to be only positive types. The ouputs of
programs that ouput mixed positive and negative types are not
observable.  Also, programs that input mixed positive and negatives
types are not executable.

@@

\noindent
with the types judgements

@@

We visually present {{eta+}} and {{eps+}} as `U'-shaped connectors. On
the left below is {{eta+}} showing the map from {{0}} to {{-b+b}} and
dualy on the right is {{eps+}} showing the map from {{-b+b}} to
0. Even though the diagrams below show a {{0}} wire for completeness,
later diagrams will always drop them in contexts where we can
implicitely introduce/eliminate then using {{zeroi}}/{{zeroe}}.

@@

The usual interpretation of the type {{b1+b2}} is that we either have
an appropriately tagged value of type {{b1}} or an appropriately
tagged value of type {{b2}}. This interpretation is mainted in the
precense of {{eta+}} and {{eps+}} in the following sense: a value of
type {{R v : -b + b}} flowing into an {{eps+}} on the lower wire
switches into a value {{L (-v): -b + b}} that flows backwards on the
upper wire. The inversion of direction is captured by the negative
sign on the value.

\paragraph*{Negative Information Flow.} 
Hence, the intuitive way to read the digrams is that they reverse the
flow of information -- in normal {{langRev}} circuits information
flows from left-to-right and {{eps+}} operations cause information to
flow from right-to-left.  The interpretation of {{eta+}} is precisely
the opposite: it turns a right-to-left flow of information in a
cicruit into a left-to-right flow.

@@

\paragraph*{Time Travelling.}
A time travelling intuition is also applicable here: if view the
left-to-right direction of information flow as the canonical forward
direction of information, then the action of {{eps+}} allows values,
aka information particles, to flow backwards in time. Particles that
flow backward in time get to see and interact with the history of
particles that they coexist with i.e. values that they exist
multiplicatively with. Hence the isomorphism {{(-b1)*b2<->-(b1*b2)}}
(whose construction is shown in Section \ref{sec:neg-constructions}).

\paragraph*{Coherence.}
The appropriate coherence condition for Compact Closed Categories,
expressed as the traingle axiom, is satisfied by {{eta+}}/{{eps+}} and
may be visualized as explained by Selinger
\cite{springerlink:10.1007/978-3-642-12821-94} as follows:

@@

\item
Type judgement.
%subcode{opsem} include main
% eta &: 0 <-> (-b) + b :& eps

@@

The following isomoprhism can be constructed similarly

@@

\paragraph*{Lifting a operation of postive types to negated types}

@@

\paragraph*{Subtraction} Wadler the reloaded paper, does not consider the
subtraction type because its ``computational interpretation is not
familiar.'' Curien and Herbelin study duality in classical logic, show that
it exchanges call-by-value and call-by-name. They extend classical natural
deduction with subtraction but give it no computational meaning. Crolard (in
the formulae-as-types interpretation of subtractive logic) address the
computational interpretation of substraction. He explains the type $A-B$ as
the of \emph{coroutines} with a local environment of type $A$ and a
continuation of type $B$. The description is complicated by what is
essentially the desire to enforce linearity constraints so that coroutines
cannot access the local environment of other coroutines. Must cite Selinger
control categories in this context of duality but I am not sure what to say:
it assumes cartesian closed categories for one thing and not symmetric
monoidal ones. Ariola, Herbelin, and Sabry also use subtractive types to
explain delimited continuations.

\paragraph*{Linear Logic} In accounts that are linear, the value and
continuation that comprise the substractive type need to be constrained to
``stay together.'' This can be achieved by various restrictions. In this work
we have no such constraints, the negative value can flow anywhere. The entire
system guarantees that any closed program would have to account for it. We
don't have to introduce special constraints to achieve that. Zeilberger in
the paper on polarity and the logic of delimited continuations uses polarized
logic: he shows that if positive and negative values are completely symmetric
except that answer types are positive, then the framework accommodates
delimited continuations. But he interprets negative values are control
operators, or as values defined by the shape of their continuations. We
simply interpret values of negative type as values flowing in the ``other''
direction.

\paragraph*{Int Construction.}
For a traced monoidal category {{C}} the Int construction produces a
Compact Closed Category called Int {{C}} \cite{joyal1996traced}.
Further we know that the target of the Int construction is isomorphic
to the target of \G construction of Abramsky \cite{Abramsky96:0} from
Haghverdi. However, note that the {{langRevEE}} is not the same as the
image of the Int construction on {{langRevT}}, since the the later
lacks a multiplicative tensor that distrbutes over the additive tensor
in Int {{langRevT}}.

@@

@@

Consider first an example in which we want to compute with the sides of a
rectangle whose area is 91 and whose length is longer than its width by 6
units. One can solve the quadratic equation to determine that the sides are 7
and 13 and proceed. This however prematuraly forces us to globally solve the
constraint. Instead we can let the two sides of the rectangle be $x$ and
$x+6$ and use the following equation to capture the desired constraint:
\[
x^2 + 6x - 91 = 0
\]
The equation introduces an isomorphism between the type $x^2 + 6x - 91$ and
the type $0$. We can now proceed to compute with the unknown $x$, being
assured that in a closed program, our computation will eventually be
consistent with the solution of the algebraic equation.

>>>>

In a technical sense, this paper extends the language of isomorphisms
{{langRev}}, with duality. Unlike Linear logic \cite{Girard87tcs} and
other systems which have one notion of duality over additive and
multiplicative components, {{langRevEE}} has two notions of duality
-- an additive duality over the monoid {{(0, +)}} and a multiplicative
duality over the monoid {{(1, *)}}. Each axis of duality also give us
a function space and hence {{langRevEE}} has an additive function
space corresponding to a notion of control or backtracking and a
multiplicative function space corresponding to a notion of unification
or constraint satisfaction.

@@

The presentation here differs from \cite{infeffects} in the sense that
we have introduced the {{0}} type here, thereby completing the monoid
{{(0,+)}} and we present a small-step operational semantics.

@@

We define observables to be only positive types. The outputs of
programs that output mixed positive and negative types are not
observable.  Also, programs that input mixed positive and negatives
types are not executable.

@@

\noindent
with the types judgments

@@

We visually present {{eta+}} and {{eps+}} as `U'-shaped connectors. On
the left below is {{eta+}} showing the map from {{0}} to {{-b+b}} and
dually on the right is {{eps+}} showing the map from {{-b+b}} to
0. Even though the diagrams below show a {{0}} wire for completeness,
later diagrams will always drop them in contexts where we can
implicitly introduce/eliminate then using {{zeroi}}/{{zeroe}}.

@@

The usual interpretation of the type {{b1+b2}} is that we either have
an appropriately tagged value of type {{b1}} or an appropriately
tagged value of type {{b2}}. This interpretation is maintained in the
presence of {{eta+}} and {{eps+}} in the following sense: a value of
type {{R v : -b + b}} flowing into an {{eps+}} on the lower wire
switches into a value {{L (-v): -b + b}} that flows backwards on the
upper wire. The inversion of direction is captured by the negative
sign on the value.

\paragraph*{Negative Information Flow.} 
Hence, the intuitive way to read the diagrams is that they reverse the
flow of information -- in normal {{langRev}} circuits information
flows from left-to-right and {{eps+}} operations cause information to
flow from right-to-left.  The interpretation of {{eta+}} is precisely
the opposite: it turns a right-to-left flow of information in a
circuit into a left-to-right flow.

@@

\paragraph*{Time Traveling.}
A time traveling intuition is also applicable here: if view the
left-to-right direction of information flow as the canonical forward
direction of information, then the action of {{eps+}} allows values,
aka information particles, to flow backwards in time. Particles that
flow backward in time get to see and interact with the history of
particles that they coexist with i.e. values that they exist
multiplicatively with. Hence the isomorphism {{(-b1)*b2<->-(b1*b2)}}
(whose construction is shown in Section \ref{sec:neg-constructions}).

\paragraph*{Coherence.}
The appropriate coherence condition for Compact Closed Categories,
expressed as the triangle axiom, is satisfied by {{eta+}}/{{eps+}} and
may be visualized as explained by Selinger
\cite{springerlink:10.1007/978-3-642-12821-94} as follows:

@@

\item
Type judgment.
%subcode{opsem} include main
% eta &: 0 <-> (-b) + b :& eps

@@

The following isomorphism can be constructed similarly

@@

\paragraph*{Lifting a operation of positive types to negated types}

@@

\paragraph*{Subtraction} Wadler the reloaded paper, does not consider the
subtraction type because its ``computational interpretation is not
familiar.'' Curien and Herbelin study duality in classical logic, show that
it exchanges call-by-value and call-by-name. They extend classical natural
deduction with subtraction but give it no computational meaning. Crolard (in
the formulae-as-types interpretation of subtractive logic) address the
computational interpretation of subtraction. He explains the type $A-B$ as
the of \emph{coroutines} with a local environment of type $A$ and a
continuation of type $B$. The description is complicated by what is
essentially the desire to enforce linearity constraints so that coroutines
cannot access the local environment of other coroutines. Must cite Selinger
control categories in this context of duality but I am not sure what to say:
it assumes cartesian closed categories for one thing and not symmetric
monoidal ones. Ariola, Herbelin, and Sabry also use subtractive types to
explain delimited continuations.

\paragraph*{Linear Logic} In accounts that are linear, the value and
continuation that comprise the subtractive type need to be constrained to
``stay together.'' This can be achieved by various restrictions. In this work
we have no such constraints, the negative value can flow anywhere. The entire
system guarantees that any closed program would have to account for it. We
don't have to introduce special constraints to achieve that. Zeilberger in
the paper on polarity and the logic of delimited continuations uses polarized
logic: he shows that if positive and negative values are completely symmetric
except that answer types are positive, then the framework accommodates
delimited continuations. But he interprets negative values are control
operators, or as values defined by the shape of their continuations. We
simply interpret values of negative type as values flowing in the ``other''
direction.

\paragraph*{Int Construction.}
For a traced monoidal category {{C}} the Int construction produces a
Compact Closed Category called Int {{C}} \cite{joyal1996traced}.
Further we know that the target of the Int construction is isomorphic
to the target of \G construction of Abramsky \cite{Abramsky96:0} from
Haghverdi. However, note that the {{langRevEE}} is not the same as the
image of the Int construction on {{langRevT}}, since the the later
lacks a multiplicative tensor that distributes over the additive tensor
in Int {{langRevT}}.

@@

Here is one way to view computations involving irrationals that might
shed some insight into how we should view recursive data types
also. For finite datatypes say of size {{n}} we can say that an
element specifies a choice of {{1/n}} and hence contains as much
information. For finite things, each value is computationally equal to
any other value. Any permutation is possible among the values. Hence
there is no metric possible on the values.  Recursive data types, such
as binary trees, have an infinite number of elements and hence the
argument that says that value specifies {{1/n}} units of information
is not longer valid. Instead to talk meaningfully about information
content, we should talk about a probability distribution on the values
of the type. However what shall we base this distribution on? Is there
a natural order beyond the unfolding order of the values of a type?

In this context, let us go back to the thought that recursive such as
{{nat=nat+1}} have no meaningful solutions, whereas trees do. But the
interesting solution of trees is precisely that they involve
irrational numbers. Our usual notation for irrationals such as
{{2^{1/2} }} can actually be viewed as a thunk representing the
infinite computation {{1.4142..}}. Irrationals in this sense have a
very different information theoretic interpretation from full real
numbers. With Chaitin's Omega numbers he postulates the existence of
real numbers that have an infinite amount of information -- ones for
example are the solution to the question about the halting properties
of every Turning machine.

Real numbers in that sense really are a series of bits of unknown
information theoretic interpretation -- a sort of {{top}}. Numbers
such as {{2^{1/2} }} are very different in the sense that even though
they contain an infinite sequence of bits, there is a finite program
that generates those bits and its only the cost of computing bits that
is infinite.  The consistency of arithmetic isn't affected by
extending the rational field by adding any specific computable
irrational. So it might be possible to design a model of computing
where we can indeed deal with them. Such a model will have to explicit
thunks for the delayed computation represented by trees and never
equate and cancel two infinite computations unless the represent
exactly the same infinite computation.

@@

Consider first an example in which we want to compute with the sides of a
rectangle whose area is 91 and whose length is longer than its width by 6
units. One can solve the quadratic equation to determine that the sides are 7
and 13 and proceed. This however prematurely forces us to globally solve the
constraint. Instead we can let the two sides of the rectangle be $x$ and
$x+6$ and use the following equation to capture the desired constraint:
\[
x^2 + 6x - 91 = 0
\]
The equation introduces an isomorphism between the type $x^2 + 6x - 91$ and
the type $0$. We can now proceed to compute with the unknown $x$, being
assured that in a closed program, our computation will eventually be
consistent with the solution of the algebraic equation.
